Component response rate variation drives stability in large complex systems

### 2018-07-28 notebook/SI.tex
Fix SI

\textbf{This supplemental information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with all of the code required to recreate the analysis in the
main text, and with additional analyses to support its conclusions. All
text, code, and data underlying this manuscript are publicly available
on \href{https://github.com/bradduthie/RandomMatrixStability}{GitHub} as
part of the RandomMatrixStability package.}

The
\href{https://github.com/bradduthie/RandomMatrixStability}{RandomMatrixStability
package} includes all functions and tools for recreating the text, this
supplemental information, and running all code; additional documentation
is also provided for functions as part of the package. The
RandomMatrixStability package is available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub}; to
download it, the
\href{https://cran.r-project.org/web/packages/devtools/index.html}{\texttt{devtools}
library} is needed.

@@

\begin{itemize}
\tightlist
\item
  \protect\hyperlink{IncrS}{Stability across increasing \(S\)}
\item
  \protect\hyperlink{ga}{Stability given targetted manipulation of
  \(\gamma\) (genetic algorithm)}
\item
  \protect\hyperlink{ecological}{Stability of ecological networks}

@@

Figure 3 of the main text reports the number of stable random complex
systems found over 1 million iterations. The table below shows the
results for all simulations of random \(M\) matrices at \(\sigma = 0.4\)
and \(C = 1\) given a range of \(S = \{2, 3, ..., 49, 50\}\). In this
table, the \texttt{A0} refers to matrices where \(\gamma = 1\), while
\texttt{A1} refers to matrices after \(Var(\gamma)\) is added and
\(\gamma \sim \mathcal{U}(0, 2)\). Each row summarises data for a given
\(S\) over 1 million randomly simulated \(M\) (\texttt{A0} and
\texttt{A1}). The column \texttt{A0\_unstable} shows the number of
\texttt{A0} matrices that are unstable, and the column
\texttt{A0\_stable} shows the number of \texttt{A0} matrices that are
stable (these two columns sum to 1 million). Similarly, the column
\texttt{A1\_unstable} shows the number of \texttt{A1} matrices that are
unstable and \texttt{A1\_stable} shows the number that are stable. The
columns \texttt{A1\_stabilised} and \texttt{A1\_destabilised} show how
many \texttt{A0} matrices were stabilised or destabilised, respectively,
by \(Var(\gamma)\).

@@

Overall, the ratio of stable \texttt{A1} matrices to stable \texttt{A0}
matrices found is greater than 1 (compare column 5 to column 3), and
this ratio increases with increasing \(S\) (column 1). Hence, more
randomly created complex systems (\(M\)) are generated given variation
in \(\gamma\) than when \(\gamma = 1\). Note that feasibility results
were ommited for the table above, but are
\protect\hyperlink{Feasibility}{reported below}.

\hypertarget{ga}{\section{\texorpdfstring{Stability given targetted
manipulation of \(\gamma\) (genetic
algorithm)}{Stability given targetted manipulation of \textbackslash{}gamma (genetic algorithm)}}\label{ga}}

Figure 4 of the main text reports the number of stable random complex
systems found over 100000 using the genetic algorithm to maximise
stability with a vector \(\mathbf{\gamma}\). Stability results for
100000 \(M\) for each \(S\) from 2-40 are shown below. Results for
\texttt{A0} indicate systems in which \(\gamma = 1\), while \texttt{A1}
refers to systems in which the genetic algorithm searched for a set of
\(\gamma\) values that stabilised the system.

@@

The distributions of nine \(\gamma\) vectors from the highest \(S\)
values are shown below. This comparison shows the high number of stable
\(M\) that can be produced through a targetted search of \(\gamma\)
values, and suggests that many otherwise unstable systems could
potentially be stabilised by an informed manipulation of their component
response times. Such a possibility might conceivably reduce the
dimensionality of problems involving stability in social-ecological or
economic systems.

@@

The distribution of \(\gamma\) values found by the genetic algorithm is
uniform. A uniform distribution was used to initialise \(\gamma\)
values, so there is therefore no evidence that a particular distribution
of \(\gamma\) is likely to be found to stabilise a matrix \(M\).

@@

While the foundational work of May (1) applies broadly to complex
networks, much attention has been given specifically to ecological
networks of interacting species. In these networks, the matrix \(M\) is
interpreted as a community matrix and each row and column is interpreted
as a single species. The effect that the density of any species \(i\)
has on the population dynamics of species \(j\) is found in \(M_{ij}\),
meaning that \(M\) holds the effects of pair-wise interactions between
\(S\) species (2, 4, 20). While May's original work (1) considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(M\) are negative), mutualist networks
(all off-diagonal elements of \(M\) are positive), and predator-prey
networks (for any pair of \(i\) and \(j\), the effect of \(i\) on \(j\)
is negative and \(j\) on \(i\) is positive, or vice versa) (2, 4, 17,
20). In general, competitor and mutualist networks tend to be unstable,
while predator-prey networks tend to be highly stabilising.

I investigated competitor, mutualist, and predator-prey networks
following Allesina et al. (2). To create these networks, I first
generated a random matrix \(M\), then changed the elements of \(M\)
accordingly. If \(M\) was a competitive network, then the sign of any
positive off-diagonal elements was reversed to be negative. If \(M\) was
a mutualist network, then the sign of any positive off-diagonal elements
was reversed to be positive. And if \(M\) was a predator-prey network,
then all \(i\) and \(j\) pairs of elements were checked; any pairs of
the same sign were changed so that one was negative and the other was
positive.

The number of stable \(M\) matrices was estimated
\protect\hyperlink{IncrS}{exactly as it was} in the main text for random
matrices for values of \(S\) from 2 to 50 (100 in the case of the
relatively more stable predator-prey interactions), except that only
100000 random \(M\) were generated instead of 1 million.

The following tables for restricted ecological communities can therefore
be compared with the random \(M\) \protect\hyperlink{IncrS}{results
above} (but note that counts from systems with comparable probabilities
of stability will be an order of magnitude lower in the tables below due
to the smaller number of \(M\) matrices generated). As with the
\protect\hyperlink{IncrS}{results above}, in the tables below,
\texttt{A0} refers to matrices when \(\gamma = 1\) and \texttt{A1}
refers to matrices after \(Var(\gamma)\) is added. The column
\texttt{A0\_unstable} shows the number of \texttt{A0} matrices that are
unstable, and the column \texttt{A0\_stable} shows the number of
\texttt{A0} matrices that are stable (these two columns sum to 100000).
Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

@@

Overall, as
expected\textsuperscript{\protect\hyperlink{ref-Allesina2012}{1}},
predator-prey communities are relatively stable while mutualist
communties are highly unstable. But interestingly, while \(Var(\gamma)\)
stabilises predator-prey and competitor communities, it does not
stabilise mutualist communities. This is unsurprising because purely
mutualist communities are characterised by a very positive (2) leading
\(\Re(\lambda)\), and it is highly unlikely that \(Var(\gamma)\) alone
will shift all real parts of eigenvalues to negative values.

@@

In the main text, for simplicity, I assumed inter-connectivity values of
\(C = 1\), meaning that all off-diagonal elements of a matrix \(M\) were
potentially nonzero and sampled from a normal distribution
\(\mathcal{N}(0, \sigma^{2})\) where \(\sigma = 0.4\). Here I present
four tables showing the number of stable communities given
\(C = \{0.3, 0. 5, 0.7, 0.9 \}\). In all cases, uniform variation in
component response time (\(\gamma \sim \mathcal{U}(0, 2)\)) led to a
higher number of stable communities than when \(\gamma\) did not vary
(\(\gamma = 1\)). In contrast to the main text, 100000 rather than 1
million \(M\) were simulated. As with the results on
\protect\hyperlink{IncrS}{stability with increasing \(S\)} shown above,
in the tables below \texttt{A0} refers to matrices when \(\gamma = 1\),
and \texttt{A1} refers to matrices after \(Var(\gamma)\) is added. The
column \texttt{A0\_unstable} shows the number of \texttt{A0} matrices
that are unstable, and the column \texttt{A0\_stable} shows the number
of \texttt{A0} matrices that are stable (these two columns sum to
100000). Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

@@

\textbf{ADD TEXT ABOUT COMPARISON WITH ALLESINA AND VALUES ACROSS SIGMA}

@@

In the main text, I considered a uniform distribution of component
response rates \(\gamma \sim \mathcal{U}(0, 2)\). The number of unstable
and stable \(M\) matrices are reported in \protect\hyperlink{IncrS}{a
table above} across different values of \(S\). Here I show complementary
results for three different distributions including an exponential,
beta, and gamma distribution of \(\gamma\) values. The shape of these
distributions is shown in the figure below.

@@

The same 100000 \(M\) matrices were used to investigate stability when
applying each of these different distributions of \(\gamma\) values. The
table below shows the number of \(M\) that were unstable (\_unst) and
stable (\_stbl) for the exponential (Exp), beta, and gamma
distributions.

@@

When feasibility was evaluated with and without variation in \(\gamma\),
there was no increase in stability for \(M\) where \(\gamma\) varied as
compared to where \(\gamma = 1\). Results below illustrate this result,
which was general to all other simulations performed.

@@

\hypertarget{refs}{}
\hypertarget{ref-Allesina2012}{}
1. Allesina, S. \& Tang, S. Stability criteria for complex ecosystems.
\emph{Nature} \textbf{483,} 205--208 (2012).

>>>>

\textbf{This supplemental information supports the manuscript
``Component response rate variation drives stability in large complex
systems'' with additional analyses to support its conclusions. All text,
code, and data underlying this manuscript are publicly available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub} as
part of the RandomMatrixStability package.}

The
\href{https://github.com/bradduthie/RandomMatrixStability}{RandomMatrixStability
package} includes all functions and tools for recreating the text, this
supplemental information, and running all code; additional documentation
is also provided for package functions. The RandomMatrixStability
package is available on
\href{https://github.com/bradduthie/RandomMatrixStability}{GitHub}; to
download it, the
\href{https://cran.r-project.org/web/packages/devtools/index.html}{\texttt{devtools}
library} is needed.

@@

\begin{itemize}
\tightlist
\item
  \protect\hyperlink{IncrS}{Stability across increasing \(S\)}
\item
  \protect\hyperlink{ga}{Stability given targeted manipulation of
  \(\gamma\) (genetic algorithm)}
\item
  \protect\hyperlink{ecological}{Stability of ecological networks}

@@

Figure 3 of the main text reports the number of stable random complex
systems found over 1 million iterations. The table below shows the
results for all simulations of random \(\mathbf{M}\) matrices at
\(\sigma = 0.4\) and \(C = 1\) given a range of
\(S = \{2, 3, ..., 49, 50\}\). In this table, the \texttt{A0} refers to
matrices where \(\gamma = 1\), while \texttt{A1} refers to matrices
after \(Var(\gamma)\) is added and \(\gamma \sim \mathcal{U}(0, 2)\).
Each row summarises data for a given \(S\) over 1 million randomly
simulated \(\mathbf{M}\) (\texttt{A0} and \texttt{A1}). The column
\texttt{A0\_unstable} shows the number of \texttt{A0} matrices that are
unstable, and the column \texttt{A0\_stable} shows the number of
\texttt{A0} matrices that are stable (these two columns sum to 1
million). Similarly, the column \texttt{A1\_unstable} shows the number
of \texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows
the number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

@@

Overall, the ratio of stable \texttt{A1} matrices to stable \texttt{A0}
matrices found is greater than 1 (compare column 5 to column 3), and
this ratio increases with increasing \(S\) (column 1). Hence, more
randomly created complex systems (\(\mathbf{M}\)) are stable given
variation in \(\gamma\) than when \(\gamma = 1\). Note that feasibility
results were ommited for the table above, but are
\protect\hyperlink{Feasibility}{reported below}.

\hypertarget{ga}{\section{\texorpdfstring{Stability given targeted
manipulation of \(\gamma\) (genetic
algorithm)}{Stability given targeted manipulation of \textbackslash{}gamma (genetic algorithm)}}\label{ga}}

Figure 4 of the main text reports the number of stable random complex
systems found over 100000 using the genetic algorithm to maximise
stability with a vector \(\mathbf{\gamma}\). Stability results for
100000 \(\mathbf{M}\) for each \(S\) from 2-40 are shown below. Results
for \texttt{A0} indicate systems in which \(\gamma = 1\), while
\texttt{A1} refers to systems in which the genetic algorithm searched
for a set of \(\gamma\) values that stabilised the system.

@@

The distributions of nine \(\gamma\) vectors from the highest \(S\)
values are shown below. This comparison shows the high number of stable
\(\mathbf{M}\) that can be produced through a targeted search of
\(\gamma\) values, and suggests that many otherwise unstable systems
could potentially be stabilised by an informed manipulation of their
component response times. Such a possibility might conceivably reduce
the dimensionality of problems involving stability in social-ecological
or economic systems.

@@

The distribution of \(\gamma\) values found by the genetic algorithm is
uniform. A uniform distribution was used to initialise \(\gamma\)
values, so there is therefore no evidence that a particular distribution
of \(\gamma\) is likely to be found to stabilise a matrix
\(\mathbf{M}\).

@@

While the foundational work of
May\textsuperscript{\protect\hyperlink{ref-May1972}{1}} applies broadly
to complex networks, much attention has been given specifically to
ecological networks of interacting species. In these networks, the
matrix \(\mathbf{M}\) is interpreted as a community matrix and each row
and column is interpreted as a single species. The effect that the
density of any species \(i\) has on the population dynamics of species
\(j\) is found in \(M_{ij}\), meaning that \(\mathbf{M}\) holds the
effects of pair-wise interactions between \(S\)
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
While May's original
work\textsuperscript{\protect\hyperlink{ref-May1972}{1}} considered only
randomly assembled communities, recent work has specifically looked at
more restricted ecological communities including competitive networks
(all off-diagonal elements of \(\mathbf{M}\) are negative), mutualist
networks (all off-diagonal elements of \(\mathbf{M}\) are positive), and
predator-prey networks (for any pair of \(i\) and \(j\), the effect of
\(i\) on \(j\) is negative and \(j\) on \(i\) is positive, or vice
versa)\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{3}}.
In general, competitor and mutualist networks tend to be unstable, while
predator-prey networks tend to be highly stabilising.

I investigated competitor, mutualist, and predator-prey networks
following Allesina et
al.\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. To create
these networks, I first generated a random matrix \(\mathbf{A}\), then
changed the elements of \(\mathbf{A}\) accordingly. If \(\mathbf{A}\)
was a competitive network, then the sign of any positive off-diagonal
elements was reversed to be negative. If \(\mathbf{A}\) was a mutualist
network, then the sign of any positive off-diagonal elements was
reversed to be positive. And if \(\mathbf{A}\) was a predator-prey
network, then all \(i\) and \(j\) pairs of elements were checked; any
pairs of the same sign were changed so that one was negative and the
other was positive.

The number of stable \(\mathbf{M = \gamma A}\) systems was estimated
\protect\hyperlink{IncrS}{exactly as it was} in the main text for random
matrices for values of \(S\) from 2 to 50 (100 in the case of the
relatively more stable predator-prey interactions), except that only
100000 random \(\mathbf{M}\) were generated instead of 1 million.

The following tables for restricted ecological communities can therefore
be compared with the random \(\mathbf{M}\)
\protect\hyperlink{IncrS}{results above} (but note that counts from
systems with comparable probabilities of stability will be an order of
magnitude lower in the tables below due to the smaller number of
\(\mathbf{M}\) matrices generated). As with the
\protect\hyperlink{IncrS}{results above}, in the tables below,
\texttt{A0} refers to matrices when \(\gamma = 1\) and \texttt{A1}
refers to matrices after \(Var(\gamma)\) is added. The column
\texttt{A0\_unstable} shows the number of \texttt{A0} matrices that are
unstable, and the column \texttt{A0\_stable} shows the number of
\texttt{A0} matrices that are stable (these two columns sum to 100000).
Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

@@

Overall, as
expected\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}},
predator-prey communities are relatively stable while mutualist
communties are highly unstable. But interestingly, while \(Var(\gamma)\)
stabilises predator-prey and competitor communities, it does not
stabilise mutualist communities. This is unsurprising because purely
mutualist communities are characterised by a very
positive\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}
leading \(\Re(\lambda)\), and it is highly unlikely that \(Var(\gamma)\)
alone will shift all real parts of eigenvalues to negative values.

@@

In the main text, for simplicity, I assumed connectance values of
\(C = 1\), meaning that all off-diagonal elements of a matrix
\(\mathbf{M}\) were potentially nonzero and sampled from a normal
distribution \(\mathcal{N}(0, \sigma^{2})\) where \(\sigma = 0.4\). Here
I present four tables showing the number of stable communities given
\(C = \{0.3, 0. 5, 0.7, 0.9 \}\). In all cases, uniform variation in
component response time (\(\gamma \sim \mathcal{U}(0, 2)\)) led to a
higher number of stable communities than when \(\gamma\) did not vary
(\(\gamma = 1\)). In contrast to the main text, 100000 rather than 1
million \(\mathbf{M}\) were simulated. As with the results on
\protect\hyperlink{IncrS}{stability with increasing \(S\)} shown above,
in the tables below \texttt{A0} refers to matrices when \(\gamma = 1\),
and \texttt{A1} refers to matrices after \(Var(\gamma)\) is added. The
column \texttt{A0\_unstable} shows the number of \texttt{A0} matrices
that are unstable, and the column \texttt{A0\_stable} shows the number
of \texttt{A0} matrices that are stable (these two columns sum to
100000). Similarly, the column \texttt{A1\_unstable} shows the number of
\texttt{A1} matrices that are unstable and \texttt{A1\_stable} shows the
number that are stable. The columns \texttt{A1\_stabilised} and
\texttt{A1\_destabilised} show how many \texttt{A0} matrices were
stabilised or destabilised, respectively, by \(Var(\gamma)\).

@@

Results below show stability results given varying interaction strengths
(\(\sigma\)) for \(C = 0.05\) (note that system size \(S\) values are
larger and increase by 10 with increasing rows). In the tables below (as
\protect\hyperlink{IncrS}{above}), \texttt{A0} and \texttt{A1} refers to
matrices for \(\gamma = 1\) and \(Var(\gamma)\), respectively.

@@

In the main text, I considered a uniform distribution of component
response rates \(\gamma \sim \mathcal{U}(0, 2)\). The number of unstable
and stable \(\mathbf{M}\) matrices are reported in
\protect\hyperlink{IncrS}{a table above} across different values of
\(S\). Here I show complementary results for three different
distributions including an exponential, beta, and gamma distribution of
\(\gamma\) values. The shape of these distributions is shown in the
figure below.

@@

The same 100000 \(\mathbf{M}\) matrices were used to investigate
stability when applying each of these different distributions of
\(\gamma\) values. The table below shows the number of \(\mathbf{M}\)
that were unstable (\_unst) and stable (\_stbl) for the exponential
(Exp), beta, and gamma distributions.

@@

When feasibility was evaluated with and without variation in \(\gamma\),
there was no increase in stability for \(\mathbf{M}\) where \(\gamma\)
varied as compared to where \(\gamma = 1\). Results below illustrate
this result, which was general to all other simulations performed.

@@

\hypertarget{refs}{}
\hypertarget{ref-May1972}{}
1. May, R. M. Will a large complex system be stable? \emph{Nature}
\textbf{238,} 413--414 (1972).

\hypertarget{ref-Allesina2012}{}
2. Allesina, S. \& Tang, S. Stability criteria for complex ecosystems.
\emph{Nature} \textbf{483,} 205--208 (2012).

\hypertarget{ref-Allesina2015}{}
3. Allesina, S. \emph{et al.} Predicting the stability of large
structured food webs. \emph{Nature Communications} \textbf{6,} 7842
(2015).
